---
title: Making a geological map with Macrostrat and ggplot
subtitle: ""
date: 2021-11-15T21:24:56.821Z
summary: Learn to use the Macrostrat API with a simple map.
draft: false
featured: false
tags:
  - R
  - figures
  - tutorial
  - teaching
image:
  filename: alabama_cretaceous_map.png
  focal_point: Smart
  preview_only: false
  alt_text: Map centered on the US State of Alabama with green outline of
    Cretaceous rocks overlain on the grey map background.
---

Geological maps are important to include as figures in many papers. Often creating them can rely on existing figures or shapefiles that are in GIS. The [Macrostrat database](https://macrostrat.org/) contains shapefiles for many Formations that can be returned using the [API](https://macrostrat.org/api/v2/geologic_units/map).

First we load the important packages. We will use data from [Natural Earth](https://www.naturalearthdata.com/) to create the basemap. We also use a few other tools to get data from [Macrostrat database](https://macrostrat.org/) and convert it to a simple feature (sf) object.


```{r Setup packages, warning=FALSE}
library(ggplot2)
library(sf)
library(ggspatial)
library(rnaturalearth)
library(rnaturalearthdata)
library(geojsonsf)
library(RCurl)
```

We load the state outlines for the map using this code.

```{r basemap}
South_base <- ne_states(country = "united states of america", returnclass = "sf")
```

Here we get data from Macrostrat. To change this to your formation of interest, use [the sift interface](https://macrostrat.org/sift/#/) to search for a unique `strat_name_id` for a formation of interest.

The data are then converted to a `wkt` and finally an `sf` object.

```{r}

y <- getURL("https://macrostrat.org/api/v2/geologic_units/map?strat_name_id=2683&format=geojson_bare",
            .opts = curlOptions(ssl.verifypeer=FALSE, verbose=TRUE))
            
sf_temp <- geojson_wkt(y)

sf_map <- st_as_sf(sf_temp, wkt = 'geometry', crs = 4326) 

```

We then use ggplot to build the map.

First, we can create just the simple background base map.

```{r}
ggplot(data = South_base) +
  geom_sf()

```
Then we can add the geological data.

```{r}
ggplot(data = South_base) +
  geom_sf()+
  geom_sf(data = sf_map,
          color = "#A6D84A",
          fill = "#A6D84A",
          alpha = .5)
```
Next we can select a smaller area and then add a North arrow.

```{r}

ggplot(data = South_base) +
  geom_sf()+
  geom_sf(data = sf_map,
          color = "#A6D84A",
          fill = "#A6D84A",
          alpha = .5)+
  coord_sf(xlim = c(-90, -82), ylim = c(28, 37.5), expand = FALSE)+
  annotation_north_arrow(location = "bl", which_north = "true", 
                         pad_x = unit(0.75, "in"), pad_y = unit(0.5, "in"),
                         style = north_arrow_fancy_orienteering)


```

Finally, we add a scale bar, background colors, dashes, and fill the ocean with a light blue.

```{r Final Map, warning=FALSE}
ggplot(data = South_base) +
  geom_sf()+
  geom_sf(data = sf_map,
          color = "#A6D84A",
          fill = "#A6D84A",
          alpha = .5)+
  coord_sf(xlim = c(-90, -82), ylim = c(28, 37.5), expand = FALSE)+
  annotation_north_arrow(location = "bl", which_north = "true", 
                         pad_x = unit(0.75, "in"), pad_y = unit(0.5, "in"),
                         style = north_arrow_fancy_orienteering)+
    annotation_scale(location = "bl", width_hint = 0.5) +
  theme(panel.grid.major = element_line(color = gray(.5),
                                        linetype = "dashed",
                                        size = 0.5),
        panel.background = element_rect(fill = "aliceblue"))+
        theme_bw()
```

And there you have a map of a geological unit overlain on a state map in R.
